ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-11-13 paper/Manuscript-Data-Package.tex
Data paper update

\title{ActiveCA: an open data product on active transportation episodes
in Canada}

@@

\begin{abstract}
This paper describes the data set of the \{ActiveCA\} R data package.
\{ActiveCA\} contains open data products to obtain impedance functions
for active transportation modes in Canada, retrieved from the General
Social Survey collections from 1986 to 2015. The package provides data
tables on walking and cycling episodes, detailing the trip origins,
destinations, and duration of each episode. The origin and destination
categories covers a wide variety of locations, such as home, work or
school, libraries, museums, restaurants, bars, sports centers, health
clinics, place of worship, and others. Addittionally, the package
details the respondent's region characteristics, specifying wheter they
live in a metropolitan area and their province of residency.
\{ActiveCA\} enables users to calculate the average travel time for each
origin-destination combination for each survey year, considering the two
active transportation modes: walking and cycling. For each Census
Metropolitan and Agglomeration Area, we estimated distance-decay curves
(impedance functions) for each destination, transportation mode and
year. The package will continue to expand with contributions from the
authors and the broader community through requests in the future.
\{ActiveCA\} is freely accessible for exploration and download from the
associated Github repository, where the documentation and code involved
in creating and manipulating data and all open data products are
detailed.
\end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This paper presents the open data product \{ActiveCA\}. Open data
products (ODPs) are the outcome of a transparent process that transforms
raw data (open or not) into analysis-ready data, in which all stages of
development follow open principles \citep{arribas-bel2021}. ODPs differ
from general open data due to their high utility, added value and open
availability. The product presented in this document is an \texttt{R}
data package that currently consists of processed data tables retrieved
from the Time Use collections of the General Social Surveys (GSS) from
1986 to 2015 \citep{statisticscanada2024}, aimed to obtain impedance
functions for active transportation modes in Canada.

To create this \texttt{R} data package, we collected, cleaned and
processed the Time Use collections from the GSS surveys to make them
ready for analysis. An \texttt{R} data package contains code, data and
documentation in a standardized format that can be installed by
\texttt{R} users via a centralized software repository, such as CRAN
(Comprehensive R Archive Network) and GitHub. Although the GSS surveys
are publicly sourced and managed by Statistic Canada, preparing them for
analysis can be time-consuming, tedious and perhaps not even possible
for those who try, due to a lack of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. \{ActiveCA\}
is freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. Although \{ActiveCA\} was
designed to obtain impedance functions, we admit and hope that its use
can be adopted in various applications that even go beyond the range of
possibilities we have imagined. Not only the data, but also all the code
documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

\hypertarget{general-social-survey-gss-collection}{%
\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}}

@@

\hypertarget{the-main-file}{%
\subsection{The Main File}\label{the-main-file}}

The Main File compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

\hypertarget{the-episode-file}{%
\subsection{The Episode File}\label{the-episode-file}}

The Episode File records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the Episode File, it is possible to link the Main
File and the Episode File by using an identifiable variable present in
both data sets.

\hypertarget{descriptive-statistics}{%
\section{Descriptive statistics}\label{descriptive-statistics}}

For each year available from the Time Use surveys, we reviewed the
episode files to identify cases with activities listed as walking or
cycling and selected the activities immediately before and after the
mobility episode. The \{ActiveCA\} package includes all 21748 episodes
that recorded walking or cycling as a mode of transportation, with trip
durations ranging from 0 to 900 minutes. Among the analyses possible
with this package, Table \ref{tab:table-01} presents descriptive
statistics on walking and cycling trips between 1986 and 2015, including
metrics such as the count of recorded trips (count), and measures of
trip duration in minutes: maximum (max), mean, median, and minimum
(min). The 1986 survey did not include bicycle trips.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item * The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(\%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination
\end{TableNotes}
\begin{longtable}[t]{ccccc>{}c|ccc>{}c|cccc}
\caption{\label{tab:table-02}\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{1986} & \multicolumn{4}{c}{1992} & \multicolumn{4}{c}{1998} \\
\cmidrule(l{3pt}r{3pt}){3-6} \cmidrule(l{3pt}r{3pt}){7-10} \cmidrule(l{3pt}r{3pt}){11-14}
\multicolumn{1}{c}{\textbf{Destination}} & \multicolumn{1}{c}{\textbf{Mode*}} & \multicolumn{1}{c}{\textbf{Min*}} & \multicolumn{1}{c}{\textbf{Med*}} & \multicolumn{1}{c}{\textbf{Max*}} & \multicolumn{1}{c}{\textbf{(\%)*}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}}\\
\midrule
Home &  & NA & NA & NA & NA & 5 & 20 & 240 & 55.6 & 2 & 15.0 & 90 & 52.9\\
Home &  & 1 & 15 & 330 & 46.4 & 1 & 10 & 300 & 59.5 & 1 & 5.0 & 255 & 51.6\\
Other's home & Cycling & NA & NA & NA & NA & 5 & 10 & 145 & 18.5 & 2 & 10.0 & 80 & 17.6\\
Other's home & Walking & 1 & 10 & 660 & 42.3 & 1 & 5 & 135 & 21.3 & 1 & 5.0 & 120 & 28.1\\
Work or school &  & NA & NA & NA & NA & 5 & 15 & 45 & 25.9 & 5 & 20.0 & 75 & 29.4\\
\addlinespace
Work or school &  & 1 & 10 & 450 & 11.3 & 2 & 10 & 60 & 19.2 & 1 & 6.5 & 75 & 20.4\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\{ActiveCA\} also enables visual analysis with traditional exploratory
data analysis techniques. Figures \ref{fig:figure-01} and
\ref{fig:figure-02} show walking and cycling trips from 1992 and 2015
through heat maps. These maps use color gradients to represent the
percentage of trips between various origins and destinations, with
darker colors indicating higher percentages and lighter colors
representing less frequent routes. To avoid overwhelming the reader, we
omitted the heat maps for the other years analyzed.

@@

\hypertarget{impedance-functions-for-canadians-metropolitan-and-census-agglomerations-areas}{%
\section{Impedance functions for Canadians Metropolitan and Census
Agglomerations
areas}\label{impedance-functions-for-canadians-metropolitan-and-census-agglomerations-areas}}

Impedance functions reveals important information about the travel
behavior of the population, by describing the relationship between the
population at an origin and their likelihood or ability to travel to
specific destinations to access opportunities \citep{soukhov2024}.

\{ActiveCA\} provides calibrated impedance functions for Canadian
Metropolitan and Census Agglomeration areas. For each combination of
year, destination, and transportation mode, we fitted the most suitable
impedance function based on empirical data from the GSS surveys.
\{ActiveCA\} includes a total of 64 distance-decay functions. These were
estimated using the \texttt{fitdistrplus} package
\citep{mullerdutang2015}, selecting the distribution with the lowest
Akaike information criterion (AIC) among exponential, gamma, log-normal,
normal, and uniform types.

Table \ref{tab:table-04} shows the best impedance functions for walking
trips to \texttt{Work\ or\ school} across survey years, where all
distributions, except for a gamma function in 1998, are log-normal.
Figure \ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others functions, these examples enable to calculate gravity-based
accessibility measures for active transportation modes across various
destinations and temporal scales in Canadian urban areas.

@@

\caption{Empirical data and impedance functions fitted for walking trips with `work or school` as destination.Walking}\label{fig:figure-03}
\end{figure}

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

This article introduces \{ActiveCA\}, an open data product in the form
of an R data package, developed after the collection, cleaning, and
processing of General Social Survey data ranging from 1986 to 2015. The
package provides analysis-ready data on active transportation episodes,
focusing on walking and cycling activities, with information on trip
origins, destinations, and duration. Additionally, the package includes
a series of impedance functions calibrated for various destinations,
considering the different transportation modes and time periods.

@@

\hypertarget{declaration-of-conflicting-interests}{%
\section{Declaration of Conflicting
Interests}\label{declaration-of-conflicting-interests}}

@@

\hypertarget{funding}{%
\section{Funding}\label{funding}}

@@

\hypertarget{orcid}{%
\section{ORCID}\label{orcid}}

@@

\hypertarget{data-availability-statement}{%
\section{Data availability
statement}\label{data-availability-statement}}

>>>>

\title{ActiveCA: Time Use Data from the General Social Survey of Canada
to Study Active Travel}

@@

\begin{abstract}
This paper describes the open data product \{ActiveCA\}. \{ActiveCA\} is
an \texttt{R} data package that contains analysis-ready data from the
General Social Survey cicles regarding active travel in Canada. The
package provides data set detailing active trip origins, destinations,
and duration, covering a wide variety of locations, such as home, work
or school, libraries, museums, restaurants, bars, sports centers, health
clinics, place of worship, and others. The package also details the
respondent's region characteristics, specifying wheter they live in a
metropolitan area and their province of residency. \{ActiveCA\} also
provides pre-estimated impedance functions for active travel for
Canadian Metropolitan Area and Census Agglomerations, which can be used
in accessibility analysis to calculate the cost of travel between
different locations. The package will continue to expand with
contributions from the authors and the broader community through
requests in the future. \{ActiveCA\} is freely accessible for
exploration and download from the associated Github repository, where
the documentation and code involved in creating and manipulating data
are detailed.
\end{abstract}

@@

\section{Introduction}\label{introduction}

This paper presents the open data product \{ActiveCA\}. Open data
products (ODPs) are the outcome of a transparent process that transforms
raw data (open or not) into analysis-ready data, in which all stages of
development follow open principles \citep{arribas-bel2021}. ODPs differ
from general open data due to their high utility, added value and open
availability.

\{ActiveCA\} is an \texttt{R} data package that provides analysis-ready
data from the General Social Survey of Canada Cycles 2 (1986), 7 (1992),
12 (1998), 19 (2005), 24 (2010), and 29 (2015)
\citep{statisticscanada2024} regarding active travel. These data can be
used to study walking and cycling in Canada for various purposes over
time, and to calculate impedance functions for accessibility analysis.
The package provides pre-estimated impedance functions for cycling and
walking trips for Census Agglomerations and Census Metropolitan Areas.

To create this \texttt{R} data package, we collected, cleaned and
processed the Time Use collections from the GSS surveys to make them
ready for analysis. An \texttt{R} data package contains code, data and
documentation in a standardized format that can be installed by
\texttt{R} users via a centralized software repository, such as CRAN
(Comprehensive R Archive Network) and GitHub. Although the GSS surveys
are publicly sourced managed by Statistic Canada
\citep{statisticscanada2024}, preparing them for analysis can be
time-consuming, tedious and perhaps not even possible for those who try,
due to a lack of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications. \{ActiveCA\}
is freely available on GitHub for all to install and freely use in the
spirit of open and reproducible research. In addition of enabling to
obtain impedance functions, \{ActiveCA\} can be adopted in various
applications that even go beyond the range of possibilities we have
imagined. Not only the data, but also all the code documenting the
processing methodology is available for consultation and evaluation in
its repository. This package contributes to reducing the barrier to
using the information contained in GSS surveys to provide data-driven
decisions in transportation analysis.

\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}

@@

\subsection{The Main File}\label{the-main-file}

The main file compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

\subsection{The Episode File}\label{the-episode-file}

The episode file records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the episode file, it is possible to link the main
file and the episode file by using an identifiable variable present in
both data sets.

\section{Methodology}\label{methodology}

For each selected cycle of the GSS surveys, we reviewed the episode
files to identify cases with activities listed as walking or cycling,
selecting the activities immediately before and after the mobility
episode. After that, we labeled the code variables with their
appropriate descriptions, identifying each origin and destination,
transportation mode, as well the province and urban classification of
the respondent's residency.

Additionally, for Census Agglomerations and Census Metropolitan Areas we
pre-estimated impedance functions for cycling and walking trips.
Impedance functions reveals important information about the travel
behavior of the population, by describing the relationship between the
population at an origin and their likelihood or ability to travel to
specific destinations to access opportunities \citep{soukhov2024}.

For each combination of year, destination, and transportation mode
(walking or cycling), we fitted the most suitable impedance function
based on empirical data from the GSS surveys. We used the
\texttt{fitdistrplus} package \citep{mullerdutang2015} to estimate the
functions, selecting the distribution with the lowest Akaike information
criterion (AIC) among exponential, gamma, log-normal, normal, and
uniform types.

\section{Results}\label{results}

\subsection{Descriptive statistics}\label{descriptive-statistics}

Considering GSS Cycles analyzed, we identified 21748 episodes that
recorded active travel episodes, with trip duration ranging from 0 to
900 minutes, to twelve different destinations. \{ActiveCA\} includes all
these episodes ready for analysis. Table \ref{tab:table-01} presents
descriptive statistics on walking and cycling trips between 1986 and
2015, including metrics such as the count of recorded trips (count), and
measures of trip duration in minutes: maximum (max), mean, median, and
minimum (min). The 1986 survey did not include bicycle trips.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item * The symbols used in this table represent the following: `Min` denotes the minimum time to reach the destination; `Max` denotes the maximum time to reach the destination; `(\%)` indicates a percentage of the total time to reach the destination; `Med` refers to the median time to reach the destination
\end{TableNotes}
\begin{longtable}[t]{ccccc>{}c|ccc>{}c|cccc}
\caption{\label{tab:table-02}\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{1986} & \multicolumn{4}{c}{1992} & \multicolumn{4}{c}{1998} \\
\cmidrule(l{3pt}r{3pt}){3-6} \cmidrule(l{3pt}r{3pt}){7-10} \cmidrule(l{3pt}r{3pt}){11-14}
\multicolumn{1}{c}{\textbf{Destination}} & \multicolumn{1}{c}{\textbf{Mode*}} & \multicolumn{1}{c}{\textbf{Min*}} & \multicolumn{1}{c}{\textbf{Med*}} & \multicolumn{1}{c}{\textbf{Max*}} & \multicolumn{1}{c}{\textbf{(\%)*}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}}\\
\midrule
Home &  & NA & NA & NA & NA & 5 & 20 & 240 & 55.6 & 2 & 15.0 & 90 & 52.9\\
Other's home & Cycling & NA & NA & NA & NA & 5 & 10 & 145 & 18.5 & 2 & 10.0 & 80 & 17.6\\
Work or school &  & NA & NA & NA & NA & 5 & 15 & 45 & 25.9 & 5 & 20.0 & 75 & 29.4\\
Home &  & 1 & 15 & 330 & 46.4 & 1 & 10 & 300 & 59.5 & 1 & 5.0 & 255 & 51.6\\
Other's home & Walking & 1 & 10 & 660 & 42.3 & 1 & 5 & 135 & 21.3 & 1 & 5.0 & 120 & 28.1\\
\addlinespace
Work or school &  & 1 & 10 & 450 & 11.3 & 2 & 10 & 60 & 19.2 & 1 & 6.5 & 75 & 20.4\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\{ActiveCA\} also enables visual analysis of active travel in Canada
using traditional exploratory data analysis techniques. Figures
\ref{fig:figure-01} and \ref{fig:figure-02} show walking and cycling
trips from 1992 and 2015 through heat maps. These maps use color
gradients to represent the percentage of trips between various origins
and destinations, with darker colors indicating higher percentages and
lighter colors representing less frequent routes. To avoid overwhelming
the reader, we omitted the heat maps for the other years analyzed.

@@

\subsection{Pre-estimated impedance functions for active travel in
Canada}\label{pre-estimated-impedance-functions-for-active-travel-in-canada}

\{ActiveCA\} includes a total of 64 pre-estimated impedance functions
for cycling and walking trips for Census Agglomerations and Census
Metropolitan Areas. Table \ref{tab:table-04} shows the best impedance
functions for walking trips to \texttt{Work\ or\ school} across survey
years, where all distributions, except for a gamma function in 1998, are
log-normal. Figure \ref{fig:figure-03} illustrates these fitted
functions (red line) alongside histograms of empirical travel times
(grey bars). As for the others pre-estimated functions, these examples
allow to calculate gravity-based accessibility measures for active
travel across various destinations and temporal scales in Canadian urban
areas.

@@

\caption{Empirical data and impedance functions fitted for walking trips with `work or school` as destination.}\label{fig:figure-03}
\end{figure}

\section{Concluding remarks}\label{concluding-remarks}

This article presents \{ActiveCA\}, an open data product that provides
analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998), 19
(2005), 24 (2010), and 29 (2015) of the GSS surveys on active travel in
Canada. In the form of an \texttt{R} data package, \{ActiveCA\} was
developed after collecting, cleaning, and processing the survey data,
providing information on origins, destinations, and duration of active
travel, as well other information. Additionally, the package includes a
series of pre-estimated impedance functions for walking and cycling
trips, considering various destinations and time periods, for Canadian
Metropolitan and Census Agglomeration areas.

@@

\section{Declaration of Conflicting
Interests}\label{declaration-of-conflicting-interests}

@@

\section{Funding}\label{funding}

@@

\section{ORCID}\label{orcid}

@@

\section{Data availability statement}\label{data-availability-statement}
